---
title: "Customize BioCarta Pathway Images"
author: "Zuguang Gu (z.gu@dkfz.de)"
date: '`r Sys.Date()`'
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Customize BioCarta Pathway Images}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, eval = TRUE, echo = FALSE}
library(knitr)
knitr::opts_chunk$set(
    fig.width = 7,
    fig.height = 7,
    error = FALSE,
    tidy  = FALSE,
    message = FALSE
)
```

BioCarta is a valuable source of biological pathways which not only provides
well manually curated pathways, but also remarkable and intuitive pathway images.
One useful features of pathway analysis which is to highlight genes of
interest on the pathway images is lost. Since the original source of
BioCarta (biocarte.com) is lost from the internet, we digged out the data from
the internet archive and formatted it into a package.

The BioCarta data is collected from
https://web.archive.org/web/20170122225118/https://cgap.nci.nih.gov/Pathways/BioCarta_Pathways.
This is an archive of BioCarta's successor website cgap.nci.nih.gov which is
also retired from internet. The snapshot is taken on 2017-01-22. Member genes as well
as the coordinates on the image are parsed from the HTML code. The script is
also shipped in the package:

```{r}
system.file("script", "process.R", package = "BioCartaImage")
```

The total pathways in the BioCarta database:

```{r}
library(BioCartaImage)
ap = all_pathways()
length(ap)
head(ap)
```

A single pathway can be obtained by providing the pathway ID:

```{r}
p = get_pathway("h_RELAPathway")
p
```

The pathway object `p` is actually a very simply list which contains
coordinates of member genes.

```{r}
str(p)
```

Similar as many other packages which contain BioCarta gene sets, the member
genes of a pathway can be obtained by `genes_in_pathway()`. You can provide
the pathway ID or the pathway object. The EntreZ ID is used as the gene ID
type.

```{r}
genes_in_pathway("h_RELAPathway")
genes_in_pathway(p)
```

Next, let's move to the main functionality of this package: customizing the
pathway.

First, as many other **grid** plotting functions, `grid.biocarta()` draws a
pathway (where the pathway image is a `raster` object internally).

```{r}
library(grid)
grid.newpage()
grid.biocarta("h_RELAPathway", color = c("1387" = "yellow"))
```

You can specify the location and how the image is aligned to the anchor point.


```{r}
grid.newpage()
grid.biocarta("h_RELAPathway", color = c("1387" = "yellow"),
    width = unit(6, "cm"))
```

You can also first create a viewport, then draw the pathway inside it.


```{r}
grid.newpage()
pushViewport(viewport(width = 0.7, height = 0.5))
grid.rect() # to show the area of the viewport
grid.biocarta("h_RELAPathway", color = c("1387" = "yellow"))
```

As the aspect ratio of the image is fixed, you can either set `width` or
`height`. If both are set, the size of the image is internally adjusted to let
the image maximally fill the plotting region.

One of the main use of the pathway image is to highlight genes of interest.
The simple use is to set the `color` argument which is a named vector where
gene EntreZ ID are names. When the colors are set, the genes are highlighted
with dashed colored borders.

```{r}
grid.newpage()
grid.biocarta("h_RELAPathway", color = c("1387" = "yellow"))
```

As normally BioCarta pathway images are colorful, it is quite difficult to
find a proper color to be distinguished from other genes. There is a more
flexible way in the package which allows to add self-defined graphics besides
the genes.

To edit the pathway image, we create the pathway grob first ("grob" is short for "graphic object").

```{r}
grob = biocartaGrob("h_RELAPathway")
```

The object `grob` basically contains a viewport and a raster image. Later we
can add graphics for single genes to it.

Graphics for single genes are added by the function `mark_gene()`. You need to
provide the pathway grob, the gene EntreZ ID and a self-defined graphics
function. As you can imagine, the input of the function is the coordinate of
the polygon of the gene in forms of two vectors: the x-coordinates and the
y-coordinates.

There are two ways to implement the graphics function. First, the function
directly returns a grob object. Later this grob is inserted to the global
pathway grob.

There is a helper function `pos_by_polygon()` which returns the position of a
certain side of the polygon.

In the following code, we add a yellow point to the left side of gene "1387"
(CBP in the image).

The graphics are drawn in the pathway image viewport which already has a
coordinate system associated. the "xscale" and "yscale" correspond to the
numbers of pixels horizontally and vertically. So `unit(1, "native")` means 1
pixel in the original image.

```{r}
grid.newpage()
grob2 = mark_gene(grob, "1387", function(x, y) {
    pos = pos_by_polygon(x, y, where = "left")
    pointsGrob(pos[1], pos[2], default.units = "native",
        pch = 16, gp = gpar(col = "yellow"))
})
grid.draw(grob2)
```

If you have complicated graphics, you can consider to use `gTree()` and
`gList()` to combine them.

If you are not familiar with `gTree()` and `gList()` or `*Grob()` functions.
You can directly use the grid plotting functions such as `grid.points()` or
`grid.lines()`. In this case, you have to set `capture` to `TRUE`, then the
graphics will be captured as grobs internally.


```{r}
grid.newpage()
grob3 = mark_gene(grob, "1387", function(x, y) {
    pos = pos_by_polygon(x, y, where = "left")
    grid.points(pos[1], pos[2], default.units = "native",
        pch = 16, gp = gpar(col = "yellow"))
}, capture = TRUE)
grid.draw(grob3)
```

With this functionality, you can implement complicated graphics to associate
to a gene. In the following example, we create a viewport and put it on the
left of the gene.

```{r}
grid.newpage()
grob4 = mark_gene(grob, "1387", function(x, y) {
    pos = pos_by_polygon(x, y)
    pushViewport(viewport(x = pos[1] - 10, y = pos[2], 
        width = unit(4, "cm"), height = unit(4, "cm"), 
        default.units = "native", just = "right"))
    grid.rect(gp = gpar(fill = "red"))
    grid.text("add whatever\nyou want here")
    popViewport()
}, capture = TRUE)
grid.draw(grob4)
```


## Session info

```{r}
sessionInfo()
```
